---
title: "21_RiverMonitoring"
author: "CBG"
date: "25 4 2022"
output: html_document
---

```{r setup, include=FALSE, results = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Describe the data treatment and plotting procedure 
The following types of plots will be made

## Data aquisition
The 2020 data has been collected from various sources, led by Øyvind Kaste. Most data was extracted from Resa, but be careful about re-analysis and quality control. The historic data used for comparison in plots is put together manually using excel. THg data with "new" sensitive NIVA method was provided by Tina. _This part should be improved_

*Installation of packages* 
```{r}
#install.packages(c("dplyr", "ggplot2", "plyr", "Rmisc", "tidyverse", "lubridate", "stringr"))
Packages <- c("plyr", "dplyr", "ggplot2",  "Rmisc", "tidyverse", "lubridate", "stringr")
lapply(Packages, library, character.only = TRUE)
```

*Functions* 
The following function will replace all "<LOD"
```{r echo = T, results = 'hide'}
dLimitx <- function(v){
  v <- as.character(v)
  isLimit <- grepl("<", v)  #grepl() is a built-in R function looks for a smaller string of characters
  n <- (sum(isLimit)/(sum(!isLimit)+sum(isLimit)))*100
  v[isLimit] <- as.numeric(gsub("<(.*)", "\\1", v[isLimit]))*(100-n)/100 #gsub(pattern, replacement, x) for substitution
  as.numeric(v)
  }
```


*Data files*
- 2020 major variables
- 2020 THg
- 2020 UV-abs indexes
- Historic data major variables
- Historic data Ca
- Historic data THg
- Historic data UV-abs indexes
- Qs data

## General start-procedure
- load required packages
- load datafiles
- check stationnames (can be misspelled from excel)
- deal with < lod observations
- Calculate nutrient fractions
- Make summary tables

### 1) Pre-treatment of data
  - for some variables, separate facet is needed when rivers differs in variable level
  - for some variables, individual adjustments is needed e.g. for optimal y-axis

Load the two datafiles of current year and previous years data 
```{r echo = T, results = 'hide'}
five <- read.table("2015_2019_data.txt", header=TRUE, sep="\t", na.string=c(""))
a2020 <- read.table("CBG3_Aquamonitor_Hovedelver_2020.txt", header=TRUE, sep="\t", na.string=c(""))
```

Fix names of stations that have been misspelled by excel formatting

```{r echo = T, results = 'hide'}
names(a2020)[5] <- "Station.name"
a2020$Station.name <- revalue(a2020$Station.name, c("M?selva v/gml E6-brua" = "Målselva", "Nidelva(Tr.heim)" = "Nidelva", "	
Numedalsl?en"="Numedalslågen","Numedalsl?en"="Numedalslågen", "Veg?dselva"="Storelva", "Vosso(Bolstadelvi)"= "Vosso", "Tanaelva"="Tana"))

five$Station.name <- revalue(five$Station.name, c("Glomma ved Sarpsfoss"="Glomma","Numedalsl†gen"="Numedalslågen","Numedalsl?gen"="Numedalslågen",  "Skienselva"="Skienselva", "Veg†rdselva" ="Storelva", "Vosso (Bolstadelvi)"="Vosso", "Nidelva (Tr.heim)"="Nidelva", "M†lselva v/gml E6-brua"="Målselva", "Tanaelva"="Tana", "M†lselv" = "Målselva", "Veg?rdselva" = "Storelva", "M?lselva v/gml E6-brua"="Målselva"))
```
Running the LOD function
```{r echo = T, results = 'hide'}
Dxa2020y <- cbind(a2020[5], apply(a2020[,-c(1:8), drop=F],2, dLimitx))
Dxfivey <- cbind(five[3:4], apply(five[,-c(1:6), drop=F],2, dLimitx))
```
Replacing columnnames for 2020 data (from Aquamonitor) to match 5-year data
```{r}
# Convert Si to SiO2 in 2020-data
Dxa2020y$SiO2 <- Dxa2020y$Si*(60.0835/28.0855)
# Change names for unity
names(Dxa2020y)[names(Dxa2020y) == 'Turbiditet'] <- 'TURB860'
names(Dxa2020y)[names(Dxa2020y) == 'TSM'] <- 'SPM'
```

### 1) Bargraphs, dodged of main chemical variables. 2020-annual average for each river including stdev error bars together with previous maximum 5-year for comparison.
First, make annual averages of the 5-year mean data. Or else the error bars in the figures will be wrong.
NOE RART HAR SKJEDD MED DATA FRA MÅLSELVA
```{r Make annual averages of 5-year data}
Dxfivey$Dates <- as.Date(Dxfivey$Date,format = "%d.%m.%Y")
Dxfivey$Year <- format(Dxfivey$Dates, format="%Y")

#MAKE summary tables of parameters, remember to expand the variable selection!
#First, summary per year
fo <- Dxfivey %>%
  group_by(Station.name, Year) %>%
  select(pH:Cr) %>% # select variables to summarise, from and to
  summarise(across(everything(), .f = list(mean = mean), na.rm = TRUE))

#then, summary together, 5-year with sd
fox<- fo %>%
  group_by(Station.name) %>%
  select(pH_mean:Cr_mean)%>%
  summarise(across(everything(), .f = list(mean = mean, sd=sd), na.rm = TRUE))

#Reshape summary table
foxy <- fox %>%
  pivot_longer(
    cols = !c(Station.name), #specify columns we want to pivot
    names_sep = "_",
    names_to  = c("chem","mu", ".value"))
foxy <- foxy[-(3)]
names(foxy) <- c("Station.name", "chem", "xmean", "xsd")
foxy$year <- "5-year mean"
foxy = foxy %>% relocate(year, .after = "Station.name")
```

Summarising the 2020 data, why not possible to add n?
```{r}
dx20 <- Dxa2020y %>%
  group_by(Station.name) %>%
  select(pH:TURB860, SiO2) %>% # select variables to summarise, from and to
  summarise(across(everything(), .f = list(mean = mean, sd=sd), na.rm = TRUE))

#Reshape summary table
df.sum20 <- dx20 %>%
  pivot_longer(
    cols = !c(Station.name), #specify columns we want to pivot
    names_sep = "_",
    names_to  = c("chem", ".value"))

df.sum20$year <- "2020 mean"

names(df.sum20) <- c("Station.name", "chem", "xmean", "xsd", "year")
df.sum20 = df.sum20 %>% relocate(year, .after = "Station.name")
```

Add the two together and make new summary table. select only those parameters to be used for the automatic plotting
```{r}
# 6) Merge the two datasets
Ni3 <- rbind(df.sum20, foxy)

dfsum2 = Ni3 %>% filter(str_detect(chem,"Ca|TURB860|SiO2|TOC|TOTN|As|Cd|Cu|Zn|Cr"))
```



Use the following for making automatised figures with correct axis-titles. BUT make sure to only include those variables you will plot. 

```{r}
#try TO MAKE NICER TITLES
dfsum2 <- Ni3 %>%
  mutate(title = case_when(
    chem== "TURB860" ~ "Turbidity",
    chem== "SPM" ~ "Suspended particulate matter",
    chem== "TOC" ~ "Total organic carbon",
    chem== "TOTN" ~ "Total nitrogen",
    chem== "SiO2" ~ "Silica",
    chem== "As" ~ "Arsenic",
    chem== "Pb" ~ "Lead",
    chem== "Cd" ~ "Cadmium",
    chem== "Cu" ~ "Copper",
    chem== "Zn" ~ "Zinc",
    chem== "Cr" ~ "Chromium",
)) %>%
  mutate(abbr = case_when(
    chem== "TURB860" ~ "TURB (FNU)",
    chem== "SPM" ~ "SPM (mg/L)",
    chem== "TOC" ~ "TOC (mg/L)",
    chem== "TOTN" ~ "TOTN (µg/L)",
    chem== "SiO2" ~ "SiO2 (mg/L)",
    chem== "As" ~ "As (µg/L)",
    chem== "Pb" ~ "Pb (µg/L)",
    chem== "Cd" ~ "Cd (µg/L)",
    chem== "Cu" ~ "Cu (µg/L)",
    chem== "Zn" ~ "Zn (µg/L)",
    chem== "Cr" ~ "Cr (µg/L)",
  ))
```

Set order of station names and legend titles
```{r}
dfsum3x <- dfsum2 %>% 
  group_by(chem, year) %>% 
  mutate(Station.name = factor(Station.name, # put education levels in logical order
                               levels = c("Glomma", "Alna", "Drammenselva", "Numedalslågen", "Skienselva",
                                         "Storelva", "Otra", "Bjerkreimselva", "Orreelva", "Vikedalselva",
                                          "Vosso", "Nausta", "Driva", "Orkla", "Nidelva", "Vefsna", "Målselva",
                                          "Altaelva", "Tana", "Pasvikelva")))



#title, abbr
dfsum3 <- dfsum3x %>% 
  group_by(chem, year) %>% 
           mutate(year=factor(year,
                     levels=c("5-year mean", "2020 mean")))
```
 
Automatised plots for variables not requiering individual adaptations such as e.g. facets
The plots shall be stored as png files in designated folder.

```{r}
plot.dfsum3 <- function(v=dfsum3){ 
  chems = unique(dfsum3$chem)
  abbrs = unique(dfsum3$abbr)
  titles = unique(dfsum3$title)
}

chems = unique(dfsum3$chem)
abbrs = unique(dfsum3$abbr)
titles = unique(dfsum3$title)

# a vector of names or regions to loop over 
for (i in seq_along(chems)){

  #a loop to produce gglopt2 graphs
  ploy <- dfsum3 %>%
    ggplot(aes(x = Station.name, fill = as.factor(year))) +
    geom_col(data = filter(dfsum3, chem == chems[i]),
             width=0.8, position=position_dodge(),
             aes(y = xmean))+
    geom_errorbar(data = filter(dfsum3, chem == chems[i]),
                  aes(x=Station.name, ymin=pmax(xmean-xsd, 0), ymax=xmean+xsd), #pmax+++ added to not show negative sd
                  width=.4,size=0.4, position=position_dodge(width=0.8), col="black")+
    scale_fill_manual(name="year", 
                      values=c("5-year mean" = "orange",
                               "2020 mean" = "dodgerblue2"))+
    theme_light()+
    theme(axis.text.y = element_text(size= 16, colour="black"),
          axis.title.x = element_blank(),
          axis.title.y = element_text(size=20, margin=margin(0,20,0,0)),
          axis.text.x = element_text(size = 16, angle=45, hjust=1,  colour="black"),
          legend.title = element_blank(),
          legend.text=element_text(size=16),
          legend.spacing.x = unit(0.2, 'cm'),
          plot.title = element_text(size = 24, hjust = 0.5),
          panel.grid.major.x = element_blank(),
          strip.background = element_blank(),
          strip.text.x = element_blank())+
    theme(legend.position="top")+
    labs(title = titles[i], y=abbrs[i])
  
print(ploy)

if (dir.exists("output7")) {  
} else {dir.create("output7")
  }

ggsave(filename = paste0("output7/",
                         chems[i],
                         "_plot.png"),
       plot = ploy,
       width = 11, height = 8.5, units = "in")

}


```


- faceting: using letters "Y" and "N" to separets Station.names to go in facet or no

```{r}
#To make facet for those of separate y-axis
dfsum3$Facet <- ifelse(dfsum3$Station.name == "Alna", 'Y',
                       ifelse(dfsum3$Station.name == "Orreelva", 'Y', "N"))
```

The plots are made manually since there are not that many plots and since different Station.names will be faceted.
In the ggplot you need to specify the variable and the axis texts. 
This option can also be used for pH which normally require the y-axis to not start at zero. Remove facet in ggplot for pH. 

```{r}
#To make plots individually to individual optimization
#Function for å definere antall siffer i akse
fmt_dcimals <- function(decimals=1){
  function(x) format(x,nsmall = decimals,scientific = FALSE)
}


plot20 <- ggplot(subset(dfsum3, chem %in% c("Ni")) , aes(x=Station.name, y=xmean, fill=year))  +
  geom_col( width=0.8, position=position_dodge())+
  geom_errorbar(aes(x=Station.name, ymin=pmax(xmean-xsd, 0), ymax=xmean+xsd),
                width=.4,size=0.4, position=position_dodge(width=0.8), col="black")+
  theme_light()+
  scale_fill_manual(values = c("orange", "dodgerblue2"))+
  theme(axis.text.y = element_text(size= 16, colour="black"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=20, margin=margin(0,20,0,0)),
        axis.text.x = element_text(size = 16, angle=45, hjust=1,  colour="black"),
        legend.title = element_blank(),
        legend.text=element_text(size=16),
        legend.spacing.x = unit(0.2, 'cm'),
        plot.title = element_text(size = 24, hjust = 0.5),
        panel.grid.major.x = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank())+
  theme(legend.position="top")+
  labs(title = "Nickel", x = "", y = "Ni (µg/L)")
  facet_wrap(.~Facet, scales="free")
```

Some trixing needs to be done to make the facets have the approperiate width

```{r}
gp <- ggplotGrob(plot20)
# optional: take a look at the grob object's layout
gtable::gtable_show_layout(gp)

# get gtable columns corresponding to the facets (5 & 9, in this case)
facet.columns <- gp$layout$l[grepl("panel", gp$layout$name)]
# get the number of unique x-axis values per facet (1 & 3, in this case)
x.var <- sapply(ggplot_build(plot20)$layout$panel_scales_x,
                function(l) length(l$range$range))

# change the relative widths of the facet columns based on
# how many unique x-axis values are in each facet
gp$widths[facet.columns] <- gp$widths[facet.columns] * x.var

# plot result
plppy = grid::grid.draw(gp)

print(plppy)
print(gp)


ggsave(filename = "Ni_20_Facet.png",
       plot = plot20,
       width = 11, height = 8.5, units = "in")

```

Some plotting variables that might be needed: 

- Subscript in axis title: "y = expression(paste("TOTP (",mu,"g/L)")"
- scaling of y-axis: "coord_cartesian(ylim = c(0.0, 0.18))" and/or "scale_y_continuous(breaks=c(0,0.03, 0.06, 0.09, 0.12, 0.15, 0.18), labels = fmt_dcimals(1))" 


The water discharge data is special since it is the sum. 
Upload manually prepared datafiles
Variables requiring individual adaptations for plotting
- water discharge needs to be summed and not averaged. 
1) sum the values of 2020 Qs
2) make average of the 5-year mean
df5 contains Qs from the previous five years
```{r Discharge}
Qs_20 <- read.table("2020_Qs.txt", header=TRUE, sep="\t", na.string=c(""))
```

```{r Naming}
Qs_20$Station.name <- revalue(Qs_20$Station.name, c("Numedalsl†gen"="Numedalslågen", "Glomma ved Sarpsfoss"="Glomma", "Vosso (Bolstadelvi)" ="Vosso", "Tanaelva"="Tana", "Nidelva (Tr.heim)" ="Nidelva", "M†lselva v/gml E6-brua"="Målselva", "	Veg†rdselva"="Storelva"))
                                                  
```
Prepare the two dataframes with equal columns and columnnames to be merged

```{r}
Qs5 <- df5[, c(1, 2, 9)]
Emp %>% group_by(Group) %>% summarise_each(funs(sum))
Qs20 <- Qs_20[, c(3:5)]
names(Qs20) <- c("Station.name", "year", "Qs")

Nix <- rbind(Qs20, Qs5)
#Ni = Ni3 %>% relocate(year, .after = "Station.name")
```


Make summary table of 5-y data

```{r}
#MAKE summary tables of parameters, remember to expand the variable selection!
df.sum <- Nix %>%
  group_by(Station.name, year) %>%
  select(Qs) %>% # select variables to summarise, from and to
  summarise(across(everything(), .f = list(mean = mean, sd = sd), na.rm = TRUE))
head(df.sum)
#Reshape summary table

df.sum2 <- df.sum %>%
  pivot_longer(
    cols = !c(Station.name, year), #specify columns we want to pivot
    names_sep = "_",
    names_to  = c("chem", ".value"))
    
#PLOTTING: different data with different types of plots
names(df.sum2) <- c("Station.name", "Year", "chem", "DisArea.mm", "xsd")
```

```{r Plotting Qs}
ggplot(data=Qs_5, aes(x=Station.name, y=DisArea.mm, fill=Year))  +
  geom_col(width=0.8, position=position_dodge())

  geom_errorbar(aes(x=Station.name, ymin=pmax(xmean-xsd, 0), ymax=xmean+xsd),
                width=.4,size=0.4, position=position_dodge(width=0.8), col="black")
  geom_col(data=df.sum2, aes(x=Station.name, y=DisArea.mm, fill=Year), width=0.8, position=position_dodge())+
  theme_light()+
  scale_fill_manual(values = c("orange", "dodgerblue2"))+
  theme(axis.text.y = element_text(size= 16, colour="black"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=20, margin=margin(0,20,0,0)),
        axis.text.x = element_text(size = 16, angle=45, hjust=1,  colour="black"),
        legend.title = element_blank(),
        legend.text=element_text(size=16),
        legend.spacing.x = unit(0.2, 'cm'),
        plot.title = element_text(size = 24, hjust = 0.5),
        panel.grid.major.x = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank())+
  theme(legend.position="top")+
  labs(title = "Nickel", x = "", y = "Ni (µg/L)")
```


### 2) Bargraphs, stacked. This is typically for fractions, e.g. nutrients, DOC/POC, etc. 
# Calculate for stacked bars, 2020-data only 

Variables: TOC+DOC, 

  - two or three components stacked
  - for some variables, separate facet is needed when rivers differ

Calculate the fractions that will be plotted. For the 2020 data only. 

```{r}
Dxa2020y$POC = Dxa2020y$TOC - Dxa2020y$DOC
Dxa2020y$TPP = Dxa2020y$TOTP - Dxa2020y$Pb.filt
Dxa2020y$Org.P = Dxa2020y$TOTP - Dxa2020y$PO4.P
Dxa2020y$Org.N = Dxa2020y$TOTN - Dxa2020y$NO3.N - Dxa2020y$NH4.N
Dxa2020y$TDN = Dxa2020y$TOTN - Dxa2020y$Nitrogen.part
```

Make new summary table

```{r}
#Negative POC values replaced by zero
Dxa2020y$POC[Dxa2020y$POC < 0] <- 0       # Set negative values to 0

head(Dxa2020y)     
#MAKE summary tables of parameters, remember to expand the variable selection!
df.sum <- Dxa2020y %>%
  group_by(Station.name) %>%
  select(TOC:TOTN, DOC:TOTP.F, TPP:TDN) %>% # select variables to summarise, from and to
  summarise(across(everything(), .f = list(mean = mean), na.rm = TRUE))

#Reshape summary table
#library(tidyverse)
df.sum2 <- df.sum %>%
  pivot_longer(
    cols = !c(Station.name), #specify columns we want to pivot
    names_sep = "_",
    names_to  = c("chem", ".value"))

#PLOTTING: different data with different types of plots
head(df.sum2)
names(df.sum2) <- c("Station.name", "chem", "xmean")

dfsum3x <- df.sum2 %>% 
  group_by(chem) %>% 
  mutate(Station.name = factor(Station.name, # put education levels in logical order
                               levels = c("Glomma", "Alna", "Drammenselva", "Numedalslågen", "Skienselva",
                                         "Storelva", "Otra", "Bjerkreimselva", "Orreelva", "Vikedalselva",
                                          "Vosso", "Nausta", "Driva", "Orkla", "Nidelva", "Vefsna", "Målselva",
                                          "Altaelva", "Tana", "Pasvikelva")))
```

Plotting of stacked will require individual adaptations using the ggplot below. Two plots need facet. 

```{r}
#To make plots individually to individual optimization
#Function for å definere antall siffer i akse
fmt_dcimals <- function(decimals=1){
  function(x) format(x,nsmall = decimals,scientific = FALSE)
}

head(dfsum3x)
#fill = forcats::fct_rev(chem)))
#PLOT stacket with possibility of facet
plotty <- ggplot(subset(dfsum3x, chem %in% c("NH4.N", "NO3.N", "Org.N")), aes(x=Station.name, y=xmean,
                                                  fill = chem))+
  geom_col(width=0.6)+
  theme_light()+
  scale_fill_manual(values = c("lightskyblue", "dodgerblue2", "dodgerblue4"), 
                    labels = c("Ammonium-N", "Nitrate-N", "Total organic-N"))+
  theme(axis.text.y = element_text(size= 16, colour="black"),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size=20, margin=margin(0,20,0,0)),
        axis.text.x = element_text(size = 16, angle=45, hjust=1,  colour="black"),
        legend.title = element_blank(),
        legend.text=element_text(size=16),
        legend.spacing.x = unit(0.2, 'cm'),
        plot.title = element_text(size = 24, hjust = 0.5),
        panel.grid.major.x = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank())+
  theme(legend.position="top")+
  labs(title = "", y =expression(paste("N (",mu,"g/L)")))+
  scale_y_continuous(breaks=c(0, 250, 500, 750, 1000, 1250, 1500)) 
  scale_y_continuous(labels = fmt_dcimals(1))
    facet_wrap(.~Facet, scales="free")
```

When using facet, the plot needs some more optimization as follows

```{r}
#Following to make facet_wrap same width  
gp <- ggplotGrob(plotty)
# optional: take a look at the grob object's layout
gtable::gtable_show_layout(gp)

# get gtable columns corresponding to the facets (5 & 9, in this case)
facet.columns <- gp$layout$l[grepl("panel", gp$layout$name)]
# get the number of unique x-axis values per facet (1 & 3, in this case)
x.var <- sapply(ggplot_build(plotty)$layout$panel_scales_x,
                function(l) length(l$range$range))

# change the relative widths of the facet columns based on
# how many unique x-axis values are in each facet
gp$widths[facet.columns] <- gp$widths[facet.columns] * x.var

# plot result
plppy = grid::grid.draw(gp)
```

saving the plot

```{r}
ggsave(filename = "OrganicP_InorganicP.png",
       plot = gp,
       width = 11, height = 8.5, units = "in")
```


  
### 3) Organic Matter quality
  - seasonal variation
  - annual averages
  - regional patterns


